---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bactcountr

<!-- badges: start -->
<!-- badges: end -->

The goal of bactcountr is to provide an automated way to calculate CFUs based on the dilutions used and to plot the final results.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("aef1004/bactcountr")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r message = FALSE}
library(bactcountr)

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)
library(readxl)

```

In order to use the first function `tidy_CFU` the data must have some columns:

- some type of naming or grouping convention to separate out the individual observations such as the columns group, mouse, organ
- dilution columns that are labeled `dilution_x` where x is the dilution used

```{r}
data("CFU_data")
head(CFU_data)
```

When the `tidy_CFU` function is used, it puts it into a tidy format. The naming/grouping columns are left alone, but the dilution and CFU columns are gathered so that each row of the dataframe represents a single observation. Any values in the original dataframe that are labeled as "TNTC" (Too Numerous To Count) are converted to NA columns as they cannot be used to calculate the CFUs.

```{r}
CFU_raw_formatted <- tidy_CFU(CFU_data)

head(CFU_raw_formatted)
```

The function `pick_one_dilution` sifts through all of the data for the groups and finds the CFU observation for each group/mouse/organ that is closest to 25 CFUs (and therefore, most likely the most accurate observation). It picks one of the dilution-CFU observations per grouping.
```{r}
CFU_one_dilution <- pick_one_dilution(CFU_raw_formatted, "CFUs", c("group", "organ", "mouse"))

head(CFU_one_dilution)
```

The `calculate_cfu` function calculates the whole CFUs and log CFUs for each observation in the data. It takes in experimental parameters such as the dilution factor used, the volume (in milliliters) used to resuspend the CFU solution, and the percent of the organ used (if organ is used, default is 1 so whole organ).
```{r}
final_data <- calculate_cfu(CFU_one_dilution, dilution_factor = 5, resuspend_volume_ml = 0.5, percent = 0.5, "dilution", "CFUs")

head(final_data)
```


## Example of full run-through of the data and plotting

This is an example reading in the data from an excel file. The functions are all run through so that the log CFUs are calculated for each group and mouse for the spleen. The results can subsequently be plotted.

```{r warning = FALSE}
# example file for the excel file
example_file_address <- system.file("extdata", "PL_D-21_CFUs.xlsx", package = "bactcountr")

# all of the functions are used to tidy the data, pick one dilution, and then calculate the log CFUs
analyzed_CFUs <- read_xlsx(example_file_address) %>%
  tidy_CFU() %>%
  pick_one_dilution("CFUs", c("group", "organ", "mouse")) %>%
  filter(organ == "Spleen") %>%
  calculate_cfu(5, 0.5, .5, "dilution", "CFUs")

# the CFUs can then be plotted
ggplot(analyzed_CFUs, aes(group, log_CFUs, color = group)) +
  geom_beeswarm(groupOnX = TRUE) +
  ggtitle("PL BCG CFUs D-21 Spleen")

```

## Alternative use without using pick_one_dilution

If you want to see the consistancy of the plating across the dilutions, you can skip the `pick_one_dilution` function and plot the calculated log_CFUs for group/mouse against the dilutions.
```{r}
analyzed_CFUs_wo_pick_one <- read_xlsx(example_file_address) %>%
  tidy_CFU() %>%
  filter(organ == "Spleen") %>%
  calculate_cfu(5, 0.5, .5, "dilution", "CFUs") %>%
  unite(col = group_mouse, group, mouse, sep = "_")

ggplot(analyzed_CFUs_wo_pick_one, aes(dilution, log_CFUs, color = group_mouse)) +
  geom_point() +
  geom_path() +
  ggtitle("PL BCG CFUs D-21 Spleen for all dilutions")

```

